Synthesizing Quotient Lenses

### 2018-03-13 papers/QuotientLenses/main.tex
fixed eval issues

\begin{document}
\title{Synthesizing Quotient Lenses}
\begin{abstract}
{\em Quotient lenses} are bidirectional transformations that respect
specified equivalence relations with the goal of 
appropriately handling inessential details in data formats. 
For example, a programmer could use a quotient lens to define 
a transformation that ignores the order of fields in XML data, so
that two XML files with the same fields but in different orders would be
considered the same. 
This paper simplifies the task of programming quotient lenses in three
steps.
First, we introduce {\em quotient regular expressions} (QREs) to
make it easy to express which portions of a data format are
inessential.  A QRE is an augmented regular expression that
simultaneously defines a regular expression and an equivalence
relation over matching strings.  Variations within the same
equivalence class are deemed inessential. 
Next, we introduce {\em QRE lenses}, which are lenses that map
between QREs.
Our key technical result is a proof that every QRE lens can be
transformed into a QRE lens that has the form $c_1 \cdot \ell \cdot
c_2$, where $c_1$ and $c_2$ are {\em canonizers} 
and $\ell$ is a bijective transformation.  Intuitively,
the canonizers normalize the data  by picking a
canonical representative of the relevant equivalence relations. 
This theorem means we can push all the normalization to the edges of a
bidirectional transformation, leaving in the middle a pure bijective
transformation.  
Finally,
we levarage this insight to {\em synthesize} QRE lenses from a pair
of QREs and example input-output pairs by reusing earlier work on
synthesizing bijective lenses.
We have implemented QREs and QRE lens synthesis 
in the bidirectional programming language Boomerang.
We evaluate the effectiveness of our approach by using it to
synthesize QRE lenses between various real-world data formats in
the {\tt data.gov} database.

@@

The following list summarizes the contributions of our work.
\begin{enumerate}
  \item We introduce {\em Quotient Regular Expressions} (QREs), which are
  a compact, convenient notation for simultaneously defining a
  regular language modulo some equivalence relation and a canonizer
  for that relation (Section~\ref{QRE}).
  \item We introduce {\em QRE lenses}, which are lenses that translate
  between data formats specified using QREs. (Section~\ref{QRE-lenses}).
  \item We descibe a particular set of QRE lens combinators and prove
  that QRE lenses defined using these combinators can be normalized
  (Theorem~\ref{normal form}), which is the main technical
  contribution of this paper.
  (Section~\ref{QRE-lenses}).
  \item We use this result to reduce the problem of {\em synthesizing} QRE
  lenses to the problem of synthesizing bijective string lenses
  previously studied in~\cite{optician}(Section~\ref{synth}).
  \item We extend Boomerang with QREs and QRE lens synthesis. 
  We use the resulting implementation to demonstrate the utility of
  our approach by synthesizing QRE lenses between
  data formats drawn from the {\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

@@

To evaluate our QRE implementation, we adapted 39 lens synthesis tasks from the
benchmark suite of the original Optician system. These benchmarks are a
combination of custom benchmarks, benchmarks derived from
FlashFill~\cite{flashfill}, and benchmarks derived from Augeas~\cite{augeas2}.
In the pre-QRE version of Optician, many of these data formats had to be
modified to work with the bijectivity constraints that Optician required. For
instance, when one format permits whitespace where the other does not allow it,
the original versions of benchmark formats were modified to allow more
whitespace, thereby restoring bijectivity (but altering the data format). With
the new QRE support, we were able to remove these alterations when porting these
benchmarks. (This experience alone suggests that QREs make the lens development
process easier.)

@@

To get a sense of how much the new expressiveness that combining QRE lenses with
synthesis saves in terms of programmer effort, we compared the amount of code
that would need to be written for each of the lens benchmarks. We use number of
nodes in the abstract syntax tree as a measure of code size, and calculated
three quantities for each lens synthesis benchmark task. We only compared the
sizes on benchmarks where the canonization functions are nontrivial.
%
\begin{itemize}
  \item[\QRESize{}] the number of AST nodes in the QRE
  specifications (including examples)\saz{CHECK THIS}\afm{no but i'll add it in}
  \item[\CanonizerAndSpecSize{}]  the number of
  AST nodes in $W(q_1)$ summed with the number of AST nodes in $\canonizer(q_1)$
  \item[\LensAndSpecSize{}] the number of AST nodes
  in $W(q_1)$ summed with the number of AST nodes in the synthesized QRE lens.
\end{itemize}

@@

Figure~\ref{fig:asts} shows the number of benchmarks that can be expressed using
at most a given number of AST nodes. For a given benchmark, we found \QRESize{}
to be about half as large as \CanonizerAndSpecSize, and about a third as large
as \LensAndSpecSize{}.\BCP{These are really the critical conclusions, and they
  are hard to see, the way the graph is drawn. I wonder whether we could instead
  present with the individual examples on the X axis (sorted by number of AST
  nodes), with a bar above each one containing three regions, colored
  differently, for LS, CS-LS, and QC-CS...}\afm{will do} This suggests that
introducing QREs saves programmers significant effort compared to both Optician
and basic Boomerang.

@@

\begin{figure}[t]
\centering
\begin{subfigure}[b]{.49\textwidth}
\centering
\includegraphics{generated-graphs/times_opt}
\caption{}
\label{subfig:lenssize}
\end{subfigure}
\begin{subfigure}[b]{.49\textwidth}
\includegraphics{generated-graphs/times_new.eps}
\caption{}
\label{subfig:examplesused}
\end{subfigure}
\caption{Runtimes of Optician and our system. In (a), we run Optician and our
  system on Optician's benchmarks. We find that there is only a negligible
  performance overhead incurred by using QREs. In (b), we run Optician with QREs
  on its benchmark suite. We find that it is able to synthesize all quotient
  lenses in under 20 seconds, and typically finished in under 5 seconds.
  \bcp{The left-hand graph, especially, is hard to make anything out of, because
    the curves are squished against the axes and have almost identical shapes.
    }\afm{will fix}}
\label{fig:times}
\end{figure}

To assess the performance impact of adding QRE support to Optician, we measured
the running time of the following configurations:
\begin{itemize}
  \item[\OpticianRuntime{}] pre-QRE Optician running on its own benchmarks.
  \item[\SystemOnOptician{}] Optician with QREs running on the pre-QRE benchmarks.
  \item[\SystemOnBenchmarks{}] Optician with QREs running on its own
  benchmarks. \BCP{What is the difference between QO and QQ??  The eight
    additional benchmarks?  Why not measure just these?  Seems like this
    graphj repeats some information from the one to the left, obscuring the
    new information.}\afm{will do}
\end{itemize}

@@

Also notable is one of the data.gov examples, which converts demographic
statistics by zip code from xml form into rdf form. Synthesizing this
transformation certainly took the longest, nearly fifteen seconds, where the
next slowest took hardly over five. To validate this performance was not due to
our overhead, we modified these format to be bijective, so the synthesis task
could be input into Optician. Optician took over 15 seconds synthesize this
modified lens. We believe this conversion took longer because the input formats
were exceptionally large, requiring conversions between over 40 fields, not due
to an overhead from synthesizing from QREs.
\section{Related Work}
\label{relwork}
This paper builds on the work of Foster et al~\cite{quotientlenses} who
introduced the theory of quotient lenses and implemented quotient lenses as a
refinement of the bidirectional string processing language
Boomerang~\cite{boomerang}. As we mentioned in
Section\cref{subsec:well-formed-qres}, all our QRE combinators can be expressed
using just the {\em normalize} combinator, which is one of the canonizer
primitives that Boomerang already supports. Also, all our QRE lens combinators
are already supported in Boomerang. Consequently Boomerang quotient lenses are
at least as expressive as our language of QRE lenses.

>>>>

\begin{document}
\title{Synthesizing Quotient Lenses}
\begin{abstract}
{\em Quotient lenses} are bidirectional transformations that respect
specified equivalence relations with the goal of 
appropriately handling inessential details in data formats. 
For example, a programmer could use a quotient lens to define 
a transformation that ignores the order of fields in XML data, so
that two XML files with the same fields but in different orders would be
considered the same. 
This paper simplifies the task of programming quotient lenses in three
steps.
First, we introduce {\em quotient regular expressions} (QREs) to
make it easy to express which portions of a data format are
inessential.  A QRE is an augmented regular expression that
simultaneously defines a regular expression and an equivalence
relation over matching strings.  Variations within the same
equivalence class are deemed inessential. 
Next, we introduce {\em QRE lenses}, which are lenses that map
between QREs.
Our key technical result is a proof that every QRE lens can be
transformed into a QRE lens that has the form $c_1 \cdot \ell \cdot
c_2$, where $c_1$ and $c_2$ are {\em canonizers} 
and $\ell$ is a bijective transformation.  Intuitively,
the canonizers normalize the data  by picking a
canonical representative of the relevant equivalence relations. 
This theorem means we can push all the normalization to the edges of a
bidirectional transformation, leaving in the middle a pure bijective
transformation.  
Finally,
we levarage this insight to {\em synthesize} QRE lenses from a pair
of QREs and example input-output pairs by reusing earlier work on
synthesizing bijective lenses.
We have implemented QREs and QRE lens synthesis 
in the bidirectional programming language Boomerang.
We evaluate the effectiveness of our approach by using it to
synthesize QRE lenses between various real-world data formats in
the Optician benchmark suite.

@@

The following list summarizes the contributions of our work.
\begin{enumerate}
  \item We introduce {\em Quotient Regular Expressions} (QREs), which are
  a compact, convenient notation for simultaneously defining a
  regular language modulo some equivalence relation and a canonizer
  for that relation (Section~\ref{QRE}).
  \item We introduce {\em QRE lenses}, which are lenses that translate
  between data formats specified using QREs. (Section~\ref{QRE-lenses}).
  \item We descibe a particular set of QRE lens combinators and prove
  that QRE lenses defined using these combinators can be normalized
  (Theorem~\ref{normal form}), which is the main technical
  contribution of this paper.
  (Section~\ref{QRE-lenses}).
  \item We use this result to reduce the problem of {\em synthesizing} QRE
  lenses to the problem of synthesizing bijective string lenses
  previously studied in~\cite{optician}(Section~\ref{synth}).
  \item We extend Boomerang with QREs and QRE lens synthesis. 
  We use the resulting implementation to demonstrate the utility of
  our approach by synthesizing QRE lenses between
  data formats drawn from the Optician benchmark suite (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

@@

To evaluate our QRE implementation, we adapted 39 lens synthesis tasks from the
benchmark suite of the original Optician system. These benchmarks are a
combination of custom benchmarks, benchmarks derived from
FlashFill~\cite{flashfill}, and benchmarks derived from Augeas~\cite{augeas2}.
In the pre-QRE version of Optician, many of these data formats had to be
modified to work with the bijectivity constraints that Optician required. For
instance, when one format permits whitespace where the other does not allow it,
the original versions of benchmark formats were modified to allow more
whitespace, thereby restoring bijectivity (but altering the data format). With
the new QRE support, we were able to remove these alterations when porting these
benchmarks. 10 of Optician's 39 benchmarks had formats modified in this way ---
introducing QREs removed these format alterations. (This experience alone
suggests that QREs make the lens development process easier.)

@@

To get a sense of how much the new expressiveness that combining QRE lenses with
synthesis saves in terms of programmer effort, we compared the amount of code
that would need to be written for each of the lens benchmarks. We use number of
nodes in the abstract syntax tree as a measure of code size, and calculated
three quantities for each lens synthesis benchmark task. We only compared the
sizes on benchmarks where the canonization functions are nontrivial.
%
\begin{itemize}
  \item[\QRESize{}] the number of AST nodes in the QRE
  specifications (including examples).
  \item[\CanonizerAndSpecSize{}]  the number of
  AST nodes in $W(q_1)$ summed with the number of AST nodes in $\canonizer(q_1)$
  and the number of examples.
  \item[\LensAndSpecSize{}] the number of AST nodes
  in $W(q_1)$ summed with the number of AST nodes in the synthesized QRE lens.
\end{itemize}

@@

Figure~\ref{fig:asts} shows the number of benchmarks that can be expressed using
at most a given number of AST nodes. On average, \CanonizerAndSpecSize{} used
44\% more AST nodes than \QRESize{}, using an average of 237 more AST nodes. On
average, \LensAndSpecSize{} used 193\% more AST nodes than \QRESize{}, using an
average of 1267 more AST nodes. This suggests that introducing QREs saves
programmers significant effort compared to both pre-QRE Optician and basic
Boomerang.

@@

\begin{figure}[t]
\centering
\begin{subfigure}[b]{.49\textwidth}
\centering
\includegraphics{generated-graphs/times_opt}
\caption{}
\label{subfig:lenssize}
\end{subfigure}
\begin{subfigure}[b]{.49\textwidth}
\includegraphics{generated-graphs/times_new.eps}
\caption{}
\label{subfig:examplesused}
\end{subfigure}
\caption{Runtimes of Optician and our system. In (a), we run Optician and our
  system on Optician's benchmarks. We find that there is only a negligible
  performance overhead incurred by using QREs. In (b), we run Optician with QREs
  on the Optician benchmarks we modified to use QREs. We find that it is able to
  synthesize all quotient lenses in under 20 seconds, and typically finished in
  under 5 seconds.
\label{fig:times}
\end{figure}

To assess the performance impact of adding QRE support to Optician, we measured
the running time of the following configurations:
\begin{itemize}
  \item[\OpticianRuntime{}] pre-QRE Optician running on its own benchmarks.
  \item[\SystemOnOptician{}] Optician with QREs running on the pre-QRE
    benchmarks.
  \item[\SystemOnBenchmarks{}] Optician with QREs running on the benchmarks that
    were altered from Optician to include quotients.
\end{itemize}

@@

\section{Related Work}
\label{relwork}
This paper builds on the work of Foster et al~\cite{quotientlenses} who
introduced the theory of quotient lenses and implemented quotient lenses as a
refinement of the bidirectional string processing language
Boomerang~\cite{boomerang}. As we mentioned in
Section\cref{subsec:well-formed-qres}, all our QRE combinators can be expressed
using just the {\em normalize} combinator, which is one of the canonizer
primitives that Boomerang already supports. Also, all our QRE lens combinators
are already supported in Boomerang. Consequently Boomerang quotient lenses are
at least as expressive as our language of QRE lenses.
